---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
options(width=80)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
library(microbenchmark)
```

### Overview

**MIIVsem** is an R package for estimating structural equation models using instrumental variables.  

Version `r utils::packageVersion("MIIVsem")` includes the following features:  

* Estimation of latent variable and simultaneous equation models.  
* Model-implied and external instrumental variable estimation.
* Equation level specification tests.
* Efficient computation from covariance matrix input.  
* Polychoric instrumental variable estimation for endogenous categorical variables.
* Impose and test within- and across-equation parameter restrictions. 
* Fast bootstrap standard errors.
* Variance and covariance parameter estimation.


### Installation

In R you can install MIIVsem from CRAN as follows:

```{r gh-installation, eval = FALSE}
install.packages("MIIVsem")
```

### Usage

MIIVsem uses a subset of the model syntax employed by 
[lavaan](http://lavaan.ugent.be/) (Rosseel, 2012) for model 
specification. The following model syntax operators are 
currently supported: 

```{r table2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
| Operators  |                                                                   | 
|------------| ----------------------------------------------------------------- |
| =~         | Used for expressing measurement relations, read as 'measured by.' |
| ~          | Used For expressing regression relations, read as 'regressed on.' |
| ~~         | For specifying variances and covariances, read as 'covaries with.'| 
| *          | For assigning equality or numerical constraints.                  | 
"
cat(tabl) 
```

### Model Syntax

**Example using Syntax Operators**

In the model below, `L1 =~ Z1 + Z2 + Z3`  indicates the 
latent variable L1 is measured by 3 indicators, `Z1`, `Z2`, and `Z3`. 
Likewise, `L2` is measured by 3 indicators, `Z4`, `Z5`, and `Z6`. The 
statement `L1 ~ L2` specifies latent  variable `L1` is regressed on latent 
variable `L2`. `Z1 ~~ Z2` indicates the error of
`Z2` is allowed to covary with the error of `Z3`. The label
`LA3` prepended to `Z3` and `Z6` in the measurement model equations 
constrains the factor loadings for `Z3` and `Z6` to equality. 


```{r}
model <- '
   L1 =~ Z1 + Z2 + LA3*Z3
   L2 =~ Z4 + Z5 + LA3*Z6
   L1  ~ L2
   Z2 ~~ Z3
'  
```

**Scaling Indicators**

Following the lavaan model syntax, latent variables are defined 
using the `=~` operator.  For first order factors, the scaling 
indicator chosen is the first observed variable on the RHS of an 
equation. For the model  below `Z1` would be chosen as the 
scaling indicator for `L1` and `Z4` would be chosen as 
the scaling indicator for `L2`. 

```{r}
model <- '
   L1 =~ Z1 + Z2 + Z3
   L2 =~ Z4 + Z5 + Z6
'
```


**Equality Constraints and Parameter Restrictions**

Within- and across-equation equality constraints on the factor loading
and regression coefficients can be imposed directly in the model syntax. 
To specify equality constraints between different parameters equivalent
labels should be prepended to the variable name using the 
`*` operator. For example, we could constrain the factor 
loadings for two non-scaling indicators of latent factor `L1` to 
equality using the following  model syntax.

```{r}
model <- '
   L1 =~ Z1 + LA2*Z2 + LA2*Z3
   L2 =~ Z4 + Z5 + Z6
'
```

Researchers also can constrain the factor loading and regression 
coefficients to specific numeric values in a similar fashion. Below 
we constrain the regression coefficient  of `L1` on `L2`
to `1`.

```{r}
model <- '
   L1 =~ Z1 + Z2 + Z3
   L2 =~ Z4 + Z5 + Z6
   L3 =~ Z7 + Z8 + Z9
   L1  ~ 1*L2 + L3
'
```

**Higher Order Factor Model**
   
In the model below, the  scaling indicator for the 
higher-order factor `H1` is taken to be `Z1`, the scaling 
indicator that would have been assigned to the first lower-order 
factor `L1`. The intercepts for lower-order latent variables 
are set to zero, by default
   
```{r}
model <- '
      H1 =~ L1 + L2 + L3
      L1 =~ Z1 + Z2 + Z3
      L2 =~ Z4 + Z5 + Z6
      L3 =~ Z7 + Z8 + Z9
   '
```

**Model Defaults**

In addition to those relationships specified in the model syntax 
MIIVsem will automatically include the intercepts of any 
observed or latent endogenous variable. The intercepts
for any scaling indicators and lower-order latent variables are
set to zero, by default. Covariances among exogenous latent
and observed  variables are included by default. 
Where appropriate the covariances of the errors
of latent and observed dependent variables are also automatically 
included in the model specification. These defaults correspond 
to those used by lavaan and `auto = TRUE`, except that 
endogenous latent variable intercepts are estimated by default, 
and the intercepts of scaling indicators are fixed to zero.

### Getting Started

**MIIV Search**

Researchers typically search for instrumental variables external to the model. 
The key property of valid instruments is that they are uncorrelated with 
equation error.  The MIIV approach proposed in Bollen (1996) finds 
instruments among observed variables already in the model. Here, the 
model specification itself implies which observed variables are uncorrelated 
with the equation disturbance. 

Using the industrialization-democracy example from Bollen (1989) we illustrate 
the MIIV Search: 

```{r, warning = FALSE, message = FALSE}
library(MIIVsem)

model <- '

    Eta1 =~ y1 + y2  + y3  + y4  
    Eta2 =~ y5 + y6  + y7  + y8    
    Xi1  =~ x1 + x2 + x3 

    Eta1 ~ Xi1  
    Eta2 ~ Xi1 
    Eta2 ~ Eta1 

    y1   ~~ y5
    y2   ~~ y4
    y2   ~~ y6
    y3   ~~ y7
    y4   ~~ y8
    y6   ~~ y8 
  '
```

```{r}
miivs(model)
```

**MIIV-2SLS Estimation**

We can also estimate the industrialization-democracy model using MIIV-2SLS:

```{r}
miive(model, bollen1989a)
```

### Additional Features

**Estimation from Sample Moments**

```{r, eval = FALSE}
sample.cov  <- cov(bollen1989a)
sample.mean <- colMeans(bollen1989a)
sample.nobs <- nrow(bollen1989a)

miive(model, sample.cov = sample.cov, sample.mean = sample.mean, sample.nobs = sample.nobs)
```

**Bootstrap Standard Errors** (Version `r utils::packageVersion("MIIVsem")`)

```{r, eval = TRUE}
microbenchmark::microbenchmark(
  fit <- miive(model, bollen1989a, se = "boot", bootstrap = 100L),
  times = 100L
)
```

**Categorical Endogenous Variables (Bollen & Maydeu-Olivares (2007))**  

```{r, eval = FALSE}
model <- ' 
    female.access =~ access1 + access2 + access3 
    male.access   =~ access4 + access5 + access6 
'

miive(model, bollen1996, ordered = c("access1", "access2","access3", "access4", "access5", "access6"))
```


### Replication of Textbook Results

Following [Henningsen and Hamann (2007)](https://www.jstatsoft.org/article/view/v023i04)  we replicate textbook results from 

**Klein's Model** I (Greene, 2003, p.381)

```{r}
data("KleinI", package = "systemfit")

model <- '
  consump  ~ corpProf + corpProfLag + wages
  invest   ~ corpProf + corpProfLag + capitalLag
  privWage ~ gnp + gnpLag + trend
'

instruments <- '
  consump  ~ govExp + taxes + govWage + trend + capitalLag + corpProfLag + gnpLag
  invest   ~ govExp + taxes + govWage + trend + capitalLag + corpProfLag + gnpLag
  privWage ~ govExp + taxes + govWage + trend + capitalLag + corpProfLag + gnpLag
'

fit <- miive(model, KleinI, instruments, miiv.check = FALSE)
estimatesTable(fit)
```



